---
title: "Excess Deaths" 
output:
  html_document:
    css: style.css
---

``` {r include = FALSE} 
library(tidyverse)
library(lubridate)
library(extrafont)
library(scales)
library(ggrepel)
library(tidycensus)
library(geojsonio)
library(rgdal)
library(broom)
library(rgeos)
library(ggtext)

# ------------------------------------------------------------------------------
# SETUP
# ------------------------------------------------------------------------------

# CDC Weekly Deaths by State and Race and Ethnicity
# https://www.cdc.gov/nchs/nvss/vsrr/covid19/excess_deaths.htm
URL <- "https://data.cdc.gov/api/views/qfhf-uhaa/rows.csv" 

# Plotting parameters  
MAX_DATE <- "2021-01-01"
```

``` {r include = FALSE}
# ------------------------------------------------------------------------------
# PROCESS EXCESS DEATHS DATA 
# ------------------------------------------------------------------------------

# Load data 
raw <- read.csv(URL)

# Factor levels 
RACE_ETHNICITY_LEVELS <- c("White", "Black", "Hispanic", "Asian", "Native American", "Other")

# Pre-process raw data 
clean <- raw %>% 
  mutate(week_end = lubridate::mdy(Week.Ending.Date), 
         race_ethnicity = str_remove(Race.Ethnicity, "Non-Hispanic "), 
         race_ethnicity = case_when(
           race_ethnicity == "American Indian or Alaska Native" ~ "Native American", 
           TRUE ~ race_ethnicity), 
         race_ethnicity = fct_relevel(race_ethnicity, RACE_ETHNICITY_LEVELS)) %>% 
  filter(Time.Period == "2020", 
         Type == "Predicted (weighted)", 
         week_end < MAX_DATE) 

# Create plotting data 
plotting <- clean %>% 
  filter(Outcome %in% c("All Cause", "COVID-19")) %>% 
  rename(diff = Difference.from.2015.2019.to.2020, 
         pct_diff = Percent.Difference.from.2015.2019.to.2020) %>% 
  pivot_wider(names_from = Outcome, values_from = c(diff, pct_diff)) %>% 
  rename(state_name = Jurisdiction, 
         state_abb = State.Abbreviation, 
         diff_all = `diff_All Cause`, 
         diff_covid = `diff_COVID-19`, 
         pct_diff_all = `pct_diff_All Cause`, 
         pct_diff_covid = `pct_diff_COVID-19`) %>% 
  select(state_name, state_abb, week_end, race_ethnicity, diff_all, diff_covid, pct_diff_all, pct_diff_covid)

# Create facet labels
labels <- clean %>% 
  mutate(diff = replace_na(Difference.from.2015.2019.to.2020, 0), 
         avg = replace_na(Number.of.Deaths, 0)) %>% 
  filter(Outcome == "All Cause") %>% 
  group_by(State.Abbreviation, race_ethnicity) %>% 
  summarise(sum_difference = sum(diff), 
            sum_avg = sum(avg), 
            pct_diff = sum_difference / (sum_avg - sum_difference)) %>%
  ungroup() %>% 
  mutate(sign = ifelse(pct_diff > 0, "+", ""), 
         label = paste0(race_ethnicity, "\n", "\n", 
                        sign, comma(sum_difference, 1), " deaths", "\n", 
                        sign, percent(pct_diff, 1), " overall"), 
         label = case_when(is.na(sign) ~ paste0(race_ethnicity, "\n", "\n", "Data Unavailable"), 
                           TRUE ~ label))
 
# ------------------------------------------------------------------------------
# HELPER FUNCTIONS 
# ------------------------------------------------------------------------------

# Get facet label 
GetLabel <- function(group) {
  labels %>% 
    filter(State.Abbreviation == STATE, 
           race_ethnicity == group) %>% 
    pull(label)
}

# Get geography 
states_unique <- clean %>% 
  select(State.Abbreviation, Jurisdiction) %>% 
  distinct()

states_lookup <- states_unique$Jurisdiction
names(states_lookup) <- states_unique$State.Abbreviation

GetGeography <- function() {
  state_name <- unname(states_lookup[STATE])
  return (state_name)
}

# Get text 
GetText <- function() {
  combined_total %>% 
    filter(State.Abbreviation == STATE) %>% 
    pull(label)
}

# ------------------------------------------------------------------------------
# PROCESS CENSUS DATA 
# ------------------------------------------------------------------------------

tables <- c(
  "B01001B_001", "B01001C_001", "B01001D_001", "B01001E_001", "B01001F_001", "B01001H_001", "B01001I_001") 

acs_names <- c(
  "B01001B_001" = "Black", 
  "B01001C_001" = "Native American", 
  "B01001D_001" = "Asian",
  "B01001E_001" = "Other", 
  "B01001F_001" = "Other", 
  "B01001H_001" = "White", 
  "B01001I_001" = "Hispanic")

acs_data <- get_acs(geography = "state", variables = tables, year = 2018)

us_total <- acs_data %>% 
  group_by(variable) %>% 
  summarise(estimate = sum(estimate), 
            NAME = "United States") 

acs_clean <- acs_data %>% 
  select(NAME, variable, estimate) %>% 
  rbind(us_total) %>% 
  mutate(race_ethnicity = recode(variable, !!!acs_names)) %>% 
  group_by(NAME, race_ethnicity) %>% 
  summarise(sum_estimate = sum(estimate)) %>% 
  ungroup() %>% 
  group_by(NAME) %>% 
  mutate(total_pop = sum(sum_estimate), 
         pct_pop = sum_estimate / total_pop) %>% 
  ungroup() 

combined_by_race <- clean %>% 
  filter(Outcome == "All Cause") %>% 
  mutate(diff = replace_na(Difference.from.2015.2019.to.2020, 0)) %>% 
  group_by(Jurisdiction, race_ethnicity) %>% 
  summarise(sum_deaths = sum(diff)) %>% 
  ungroup() %>% 
  group_by(Jurisdiction) %>% 
  mutate(total_deaths = sum(sum_deaths), 
         pct_deaths = sum_deaths / total_deaths) %>% 
  ungroup() %>% 
  left_join(acs_clean, by = c("Jurisdiction" = "NAME", "race_ethnicity" = "race_ethnicity")) %>% 
  left_join(states_unique, by = "Jurisdiction") %>% 
  select(Jurisdiction, State.Abbreviation, race_ethnicity, pct_deaths, pct_pop) %>% 
  pivot_wider(names_from = race_ethnicity, values_from = c(pct_deaths, pct_pop))

combined_total <- clean %>% 
  filter(Outcome %in% c("All Cause", "COVID-19")) %>% 
  mutate(diff = replace_na(Difference.from.2015.2019.to.2020, 0)) %>% 
  group_by(State.Abbreviation, Outcome) %>% 
  summarise(sum_deaths = sum(diff)) %>% 
  pivot_wider(names_from = Outcome, values_from = sum_deaths) %>% 
  rename(deaths_all = `All Cause`, 
         deaths_covid = `COVID-19`) %>% 
  left_join(combined_by_race, by = "State.Abbreviation") %>% 
  mutate(pct_covid = deaths_covid / deaths_all, 
         label = paste0(comma(deaths_all, 1), 
                        " more people have died in 2020 relative to previous years. ", 
                        comma(deaths_covid, 1), " (or ", percent(pct_covid, 1), 
                        ") of these deaths were directly attributed to COVID-19."))
```

``` {r include = FALSE} 
# ------------------------------------------------------------------------------
# CREATE PLOTS 
# ------------------------------------------------------------------------------

CreatePlot <- function(plotting_data) {
  p <- ggplot(
    data = plotting_data %>% 
      filter(state_abb == STATE), 
    aes(x = week_end)) +
    geom_hline(yintercept = 0, size = 0.2, color = "#e3e3e3") + 
    geom_area(aes(y = diff_all), group = 1, fill = "#fec44f") +
    geom_area(aes(y = diff_covid), group = 1, fill = "#d95f0e") +
    geom_text_repel(
      data = . %>%
        group_by(race_ethnicity) %>%
        filter(!is.na(diff_all)) %>% 
        arrange(desc(diff_all, diff_covid)) %>% 
        slice(1), 
      aes(y = diff_all,  
          label = ifelse(diff_all > 0, 
                         paste0(paste0("+", comma(diff_all, 1), "\n+", percent(pct_diff_all / 100, 1))), 
                         paste0(paste0(comma(diff_all, 1), "\n", percent(pct_diff_all / 100, 1))))), 
      family = "Lato", size = 3.0, vjust = "outward", hjust = "outward") +
    facet_wrap(~ race_ethnicity, labeller = as_labeller(GetLabel), nrow = 1) + 
    scale_x_date(date_labels = "%b", date_breaks = "3 month") + 
    theme_minimal(base_size = 12, base_family = "Lato") + 
    theme(axis.title.x = element_blank(), 
          axis.title.y = element_blank(), 
          panel.grid.major.y = element_blank(), 
          panel.grid.minor.y = element_blank(), 
          panel.grid.minor.x = element_blank(), 
          panel.grid.major.x = element_line(size = 0.2, color = "#e3e3e3"), 
          plot.title = element_text(hjust = 0.5),
          axis.text.y = element_blank()) 
  
  return (p)
}
```

Updated: January 2021 

---

### Excess deaths provide a more complete picture of the pandemic's mortality  

In April, [the CDC began publishing data on excess deaths](https://www.cdc.gov/nchs/nvss/vsrr/covid19/excess_deaths.htm) – the number of deaths each week in 2020 compared to the expected number of deaths in previous years. These data are broken out by race and ethnicity for each state. The full methodology from the CDC's National Center for Health Statistics (NCHS) is available in [these technical notes](https://www.cdc.gov/nchs/nvss/vsrr/covid19/excess_deaths.htm#techNotes).  

By taking historical trends into account, excess deaths provide a more complete account of the true mortality from the pandemic relative to the official count of deaths directly attributed to COVID-19. Excess deaths take into account indirect deaths from the pandemic – for example, from people who were never tested for COVID-19 or from people who were unable to access medical treatment because of the pandemic's burden on health systems. 

A much higher share of excess deaths are attributed to COVID-19 in [states that have been conducting a large amount of testing](https://www.nytimes.com/interactive/2020/us/coronavirus-testing.html) throughout the pandemic – largely concentrated in the northeast. 

---

``` {r echo = FALSE, warning = FALSE, results = "asis", fig.align = "center"}
# Source: https://www.r-graph-gallery.com/328-hexbin-map-of-the-usa.html

# Load and process hexgrid geojson 
spdf <- geojson_read("data/us_states_hexgrid.geojson",  what = "sp")

spdf@data = spdf@data %>% 
  mutate(google_name = gsub(" \\(United States\\)", "", google_name))
spdf_fortified <- tidy(spdf, region = "google_name")
centers <- cbind.data.frame(data.frame(gCentroid(spdf, byid = TRUE), id = spdf@data$iso3166_2))

hex_combined <- spdf_fortified %>% 
  left_join(combined_total, by = c("id" = "Jurisdiction"))

# Plot national hex map 
hex_combined <- hex_combined %>% 
  mutate(pct_covid_bins = cut(
    pct_covid, 
    breaks = c(-Inf, 0.2, 0.4, 0.6, 0.8, Inf), 
    labels = c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%")))

ggplot() +
  geom_polygon(data = hex_combined, 
               aes(x = long, y = lat, group = group, fill = pct_covid_bins),
               color = "white") +
  geom_text(data = centers, 
            aes(x = x, y = y, label = id, family = "Lato")) +
  scale_fill_manual(
    values = c("#1d91c0", "#41b6c4","#7fcdbb", "#c7e9b4", "#ffffcc"), 
    name = "\n", 
    guide = guide_legend(
      keywidth = unit(26, units = "mm"), 
      label.position = "bottom", 
      title.position = "top", 
      nrow = 1)) + 
  theme_void(base_family = "Lato") + 
  labs(title = "What percent of each state's excess deaths were directly attributed to COVID-19?") + 
  theme(legend.position = "top", 
        plot.title = element_text(hjust = 0.5))
```

---

Relative to earlier in the pandemic, a higher share of excess deaths in the most recent wave were attributed to COVID-19. In late-March, fewer than half of all excess deaths were directly attributed to COVID-19. Even through the summer months, this share was less than 60%. By December, roughly 90% of all excess deaths were attributed to COVID-19.  

---

``` {r echo = FALSE, warning = FALSE, message = FALSE, results = "asis", fig.align = "center"}
# National aggregated plot 
STATE <<- "US"

us_plotting <- plotting %>% 
  filter(state_abb == STATE) %>% 
  mutate(diff_all = replace_na(diff_all, 0), 
         diff_covid = replace_na(diff_covid, 0)) %>% 
  group_by(state_abb, week_end) %>% 
  summarise(diff_all = sum(diff_all), 
            diff_covid = sum(diff_covid)) 

colors <- c("COVID-19 Deaths" = "#d95f0e", "All Excess Deaths" = "#fec44f")

ggplot(data = us_plotting, aes(x = week_end)) +
  geom_hline(yintercept = 0, size = 0.2, color = "#e3e3e3") + 
  geom_area(aes(y = diff_all, fill = "All Excess Deaths"), group = 1) +
  geom_area(aes(y = diff_covid, fill = "COVID-19 Deaths"), group = 1) +
  scale_x_date(date_labels = "%b", date_breaks = "3 month") + 
  geom_label(aes(x = as.Date("2020-02-27"), y = 23200, 
                 label = "There were +24,535 excess deaths\non the week ending on April 11"), 
             fill = "#fec44f", family = "Lato", size = 3, label.size = 0, label.r = unit(0.0, "lines")) + 
  geom_label(aes(x = as.Date("2020-02-25"), y = 15800, label = "There were 16,213 COVID-19\ndeaths that week"), 
             color = "white", fill = "#d95f0e", family = "Lato", size = 3, label.size = 0, label.r = unit(0.0, "lines")) + 
  geom_hline(yintercept = max(us_plotting$diff_all), linetype = "dashed", color = "#fec44f", size = 0.5) + 
  geom_hline(yintercept = max(us_plotting$diff_covid), linetype = "dashed", color = "#d95f0e", size = 0.5) + 
  theme_minimal(base_size = 12, base_family = "Lato") + 
  scale_y_continuous(label = label_number_si()) + 
  scale_fill_manual(values = colors) + 
  labs(title = "How has the share of excess deaths directly attributed to \n COVID-19 shifted over the course of the pandemic?") + 
  theme(axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid.major.y = element_blank(), 
        panel.grid.minor.y = element_blank(), 
        panel.grid.minor.x = element_blank(), 
        plot.title = element_text(hjust = 0.5), 
        legend.position = "top", 
        legend.title = element_blank(), 
        panel.grid.major.x = element_line(size = 0.2, color = "#e3e3e3"), 
        plot.title.position = "plot") 
```

---

### The pandemic continues to disproportately burden people of color 

A lower share of excess deaths were attributed to COVID-19 among Black and Hispanic communities, relative to the White population – even in the most recent wave of the pandemic. These disparities in excess deaths highlight the extent of mortality resulting from persistent and structural inequities in access to health care in the United States. 

These disparities can be seen at the national level and across states – particularly among states with large Black and Hispanic populations. 


---

``` {r echo = FALSE, warning = FALSE, fig.width = 11, fig.height = 2.5, results = "asis", fig.align = "center"}
# National plot by race/ethnicity 
cat("### ", GetGeography(), "\n")
cat(GetText(), "\n")
plot(CreatePlot(plotting))
```

---

``` {r echo = FALSE, warning = FALSE, fig.width = 11, fig.height = 2.5, results = "asis", fig.align = "center"}
#  State plots by race/ethnicity
states_loop <- states_unique %>%
  arrange(Jurisdiction) %>%
  select(State.Abbreviation) %>%
  filter(! State.Abbreviation == "US") %>%
  pull()

for (i in states_loop) {
  STATE <<- i
  cat("\n")
  cat("### ", GetGeography(), "\n")
  cat(GetText(), "\n")
  plot(CreatePlot(plotting))
  cat("\n")
}
```

---

### Technical notes
The charts above use excess deaths estimates based on [the CDC's predicted (weighted) methodology](https://www.cdc.gov/nchs/nvss/vsrr/covid19/excess_deaths.htm#techNotes), which accounts for under-reporting from jurisdictions in recent weeks. Only data through the end of October is included. [2018 American Community Survey 5-Year Estimates](https://data.census.gov/cedsci/table?q=B01001&g=0500000US48411&tid=ACSDT5Y2018.B01001) are used for the race and ethnicity representation in each state. 

---
